---
title: "Hirsch Index Analysis"
format: 
  html:
    code-fold: true
    embed-resources: true
execute:
  warning: false
editor: source
---

```{r}
#| label: load_libraries
#| include: FALSE
library(here)
library(tidyverse)
library(lme4)
library(texreg)
library(specr)
library(kableExtra)
library(gt)
library(ggrepel)
library(qualpalr)
options(knitr.kable.NA = "")
```

```{r}
#| label: color_palette
#| include: FALSE

# creating color palettes for twenty cases is not easy but lets see if qualpalr
# can help us
pal <- qualpal(20, "pretty_dark")
```

```{r}
#| label: read-data

scholars <- readRDS(here("data","hind19.df.rdata")) |>
  mutate(prop_sole=nps/np6019, 
         discipline=factor(discipline),
         sm_field=factor(sm_field),
         cntry=factor(cntry)) |>
  select(h19, age, proportion_female, prop_sole, discipline, sm_field, cntry, 
         unicount, sm_field_frac) |>
  rename(h_index=h19, prop_female=proportion_female, field=sm_field, 
         country=cntry, uni_pub_cnt=unicount, specialization = sm_field_frac)

```

Lets check missing values on the original data for the variables we have.

```{r}
#| label: tbl-check-missing
#| tbl-cap: Number of missing values on variables in analytical dataseet

sapply(scholars, function(x) {sum(is.na(x))}) %>%
  enframe(name = "Variable", value = "# of missing") %>%
  gt() %>%
  fmt_number(2, decimals=0)
```

We are missing almost 10K gender cases and a few cases for discipline, field, and specialization.

```{r}
#| label: tbl-drop-missing
#| tbl-cap: Sample size before and after missing values dropped

n_before <- nrow(scholars)
scholars <- scholars %>%
  drop_na()
n_after <- nrow(scholars)
tibble(Before=n_before, After=n_after) %>%
  gt()  %>%
  fmt_number(1:2, decimals=0)
```

We lose `r n_before-n_after` cases, mostly from the `prop_female` variable.

```{r}
#| label: clean_data
#| output: FALSE

# now lets add in discipline and field cluster means for prop_female and prop_sole
disciplines <- scholars |>
  group_by(discipline) |>
  summarize(disc_prop_female=mean(prop_female),
            disc_prop_sole=mean(prop_sole),
            disc_uni_pub_cnt=mean(uni_pub_cnt),
            disc_age=mean(age),
            disc_specialization=mean(specialization),
            n_disc=n())

fields <- scholars |>
  group_by(field) |>
  summarize(field_prop_female=mean(prop_female),
            field_prop_sole=mean(prop_sole),
            n_field=n())

# now merge it back together
scholars <- scholars |>
  left_join(disciplines) |>
  left_join(fields)

# lets also put in mean h_index by discipline and add back in field information
# for fun visualizations
disciplines <- scholars |>
  group_by(discipline) |>
  summarize(disc_h_index=mean(h_index),
            field=unique(field)[1]) |>
  right_join(disciplines) %>%
  left_join(fields)
```

# Between discipline data visualization

```{r}
#| label: fig-scatter_percent_female
#| fig-cap: Scatterplot of mean h-index and percent female across disciplines. Blue line is best fitting line unweighted by discipline sample size. Red line is the same but with nursing excluded.
ggplot(disciplines, aes(x=disc_prop_female, y=disc_h_index))+
  geom_point(aes(size=n_disc, color=field), alpha=0.7)+
  geom_smooth(method="lm", se=FALSE)+
  geom_smooth(data=filter(disciplines, discipline!="Nursing"), color="red",
              method="lm", se=FALSE)+
  geom_text_repel(aes(label=discipline), max.overlaps=6, size=2)+
  scale_x_continuous(labels=scales::percent)+
  scale_color_manual(values=pal$hex)+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_text(size = 6), 
        legend.text = element_text(size = 4))+
  labs(x="percent female in discipline", y="mean h-index within discipline",
       color="Field")+
  guides(size=FALSE)
```

@fig-scatter_percent_female shows a slightly negative relationship across disciplines between feminization of the discipline and mean h-index. The field of nursing is an outlier, but as the two fitted lines show, it is not an influential point.

```{r}
#| label: fig-scatter-percent_sole
#| fig-cap: Scatterplot of mean h-index and percent sole author across disciplines. Blue line shows best-fitting OLS line unweighted by discipline size.
ggplot(disciplines, aes(x=disc_prop_sole, y=disc_h_index))+
  geom_point(aes(size=n_disc, color=field), alpha=0.7)+
  geom_smooth(method="lm", se=FALSE)+
  geom_text_repel(aes(label=discipline), max.overlaps=6, size=2)+
  scale_x_continuous(labels=scales::percent)+
  scale_color_manual(values=pal$hex)+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_text(size = 6), 
        legend.text = element_text(size = 4))+
  labs(x="percent sole authored in discipline", y="mean h-index within discipline")+
  guides(size=FALSE)
```

@fig-scatter-percent_sole shows the relationship between sole authorship tendencies in a discipline and mean h-index. Disciplines with more sole authorship have a lower mean h-index and the relationship is quite a strong one. It is also somewhat curvilinear.

# Multilevel models

We now turn to multilevel models to examine both within and between discipline and field effects on h-index values. In the first set of models, we use random intercept models without covariates to partition the variance in the h-index to the individual, discipline, and field levels. We first fit two separate models that partition by discipline and field, respetively and then we do a final model that uses a three level model (individual, discipline, field) to partition the variance in discipline and field simultaneously.

```{r}
#| label: baseline models
model_base_disc <- lmer(h_index ~ 1+(1 | discipline), data=scholars)
model_base_field <- lmer(h_index ~ 1+(1 | field), data=scholars)
model_base_both <- lmer(h_index ~ 1+(1 | discipline)+(1 | discipline:field), 
                        data=scholars)

models_base <- list(model_base_disc, model_base_field, model_base_both)
```

```{r}
#| label: tbl-icc_base
#| tbl-cap: Partitioning of variance in h-index to different levels

# get icc across models in a nice tidy table
map(models_base, function(x) {
              temp <- icc_specs(x) |>
                select(grp, percent)  |>
  mutate(grp=ifelse(grp=="discipline:field", "field", grp),
         grp=factor(grp, levels=c("discipline","field","Residual"),
                    labels=c("Discipline","Field","Individual")))
            }) |>
  reduce(full_join, by="grp") |>
  arrange(grp) |>
  kbl(digits=1, col.names = c("Grouping","Model 1","Model 2","Model 3")) |>
  kable_paper()
```

@tbl-icc_base shows the percentage of the overall variation in h-index scores that can be accounted for at each level. About 68% of the variation in h-indices occurs among individuals in the same discipline, with the remaining third of the variance occurring across disciplines and fields. The results clearly show that fields do not add much beyond the variance at the discipline level. When both levels are considered, disciplinary differences account for about 28% of the variance and fields only for about 4%. 

Based on these results we used a two-level model with only disciplines as the higher level unit for our further substantive analysis. For this analysis, we turned to within-between models to understand the role of two variables both within and between disciplines. These two variables are

* gender, measured by the likelihood that a given name indicates a female gender.
* proportion of publications that are sole-authored.

The basic structure of the within-between models is:

$$y_{ij}=\beta_{0j}+\beta_1(x_{ij}-\bar{x}_j)+\beta_2(\bar{x}_j)+\upsilon_{0j}+\epsilon_{ij}$$

Where $y_{ij}$ is the h-index score of scholar $i$ in discipline $j$. The random intercept $\beta_0j$ is allowed to vary over disciplines. The variable $x$ is the focal variable of interest. It is included at both the individual level (using group mean centering) and at the discipline level, by including its mean at the discipline level. This particular model structure allows us to understand the within discipline effect and between discipline effect of $x$ simultaneously. The $\beta_1$ slope is identical to the slope one would get if they were to instead fit a traditional econometric fixed-effects model (i.e. include dummies for the disciplines), but unlike a traditional fixed-effects model, we can still estimate the between discipline part of the model. The $\beta_2$ slope tells us how much an increase in the mean of the x variable at the disciplinary changes the expected h-index score.

In practice, we include both of our key variables in the model at the same time. We also include a control variable for age which is grand mean centered. The results are shown in the table below.

```{r}
#| label: tlb-model_wb
#| tbl-cap: test
#| results: asis

# main effects are group mean centered
model_wb_disc <- lmer(h_index ~ 
                        I(prop_female-disc_prop_female)+I(100*disc_prop_female)+
                        I(100*(prop_sole-disc_prop_sole))+I(100*disc_prop_sole)+
                        (1|discipline), data=scholars)

# add a model with controls for age, specialization and the number of elite scholars per university (all grand mean centered)
model_wb_controls <- update(model_wb_disc, .~.+I(age-mean(age))+I(specialization-mean(specialization))+I(uni_pub_cnt-mean(uni_pub_cnt)))

knitreg(list(model_base_disc, model_wb_disc, model_wb_controls),
        custom.coef.names = c("Intercept",
                              "Probability [0-1] of female name (discipline mean centered)",
                              "Discipline mean percent female name",
                              "Percent sole authored publications (discipline mean centered)",
                              "Discipline mean percent sole authored publications",
                              "Age (grand mean centered)",
                              "Specialization (grand mean centered)",
                              "Highly productive university count (grand mean centered)"),
        custom.model.names = c("ANOVA Model","W-B Model","w/controls"),
        caption = "Multilevel models predicting H-index score. Clustering at the disciplinary level.",
        caption.above = TRUE)
```

The results of these models are as follows:

* Within the same discipline, women have h-indexes 2.47 points lower than men, on average. 
* Within the same discipline, a 1% percent increase in sole-authored publication is associated with a 0.47 point reduction in the h-index score.
* Between disciplines, the female concentration of the discipline had little effect on the mean h-index for the discipline. A 1% increase in the percent female is associated with a 0.08 increase in the mean h-index.
* Between disciplines, a 1% increase in the mean percent of sole authored publications is associated witha a 0.48 reduction in the mean h-index. 

So, in summary gender inequity exists within disciplines and inequity by sole authoring tendencies exists both within and between disciplines.

# Sensitivity analysis

## Grouping disciplines

```{r}
# let's remove non social science or socail science fields and repeat

disciplines2 <- disciplines %>% filter(field!="Visual & Performing Arts" , field!="Historical Studies" , field!="Philosophy & Theology")

```


# Discipline and field descriptive statistics

```{r}
#| label: tbl-discipline-stats
#| tbl-cap: Discipline specific statistics. Disciplines are sorted from highest h-index to lowest.
temp <- disciplines |>
  select(discipline, field, n_disc, disc_h_index, disc_prop_female, 
         disc_prop_sole) |>
  mutate(disc_prop_female=100*round(disc_prop_female, 3),
         disc_prop_sole=100*round(disc_prop_sole, 3)) |>
  arrange(desc(disc_h_index))

temp |>
  kbl(col.names=c("Discipline","Field","N","Mean h-index", "% female", 
                  "% sole")) |>
  kable_paper() |>
  column_spec(2, color = "white",
              background = spec_color(as.numeric(temp$field)))
```

```{r}
fields |>
  select(field, n_field, field_prop_female, field_prop_sole) |>
  mutate(field_prop_female=100*round(field_prop_female, 3),
         field_prop_sole=100*round(field_prop_sole, 3)) |>
  arrange(desc(n_field))|>
  kbl(col.names=c("Field","N", "% female", "% sole")) |>
  kable_paper() 
```

# A Series of Field Tests

Medical and Public Health Field: Clinical Medicine, Biomedical Research, Public Health & Health Services 

```{r}
#| label: tlb-model_med
#| tbl-cap: test
#| results: asis

medical <- scholars %>% filter(field=="Clinical Medicine" | field=="Biomedical Research" | field=="Public Health & Health Services" )

model_base_med <- lmer(h_index ~ 1+(1 | discipline), data=medical)

# main effects are group mean centered

model_med_disc <- lmer(h_index ~ 
                        I(prop_female-disc_prop_female)+I(100*disc_prop_female)+
                        I(100*(prop_sole-disc_prop_sole))+I(100*disc_prop_sole)+
                        (1|discipline), data=medical)

# add a model with controls for age, specialization and the number of elite scholars per university (all grand mean centered)
model_med_controls <- update(model_med_disc, .~.+I(age-mean(age))+I(specialization-mean(specialization))+I(uni_pub_cnt-mean(uni_pub_cnt)))

knitreg(list(model_base_med, model_med_disc, model_med_controls),
        custom.coef.names = c("Intercept",
                              "Probability [0-1] of female name (discipline mean centered)",
                              "Discipline mean percent female name",
                              "Percent sole authored publications (discipline mean centered)",
                              "Discipline mean percent sole authored publications",
                              "Age (grand mean centered)",
                              "Specialization (grand mean centered)",
                              "Highly productive university count (grand mean centered)"),
        custom.model.names = c("ANOVA Model","W-B Model","w/controls"),
        caption = "Multilevel models predicting H-index score for Medical Field. Clustering at the disciplinary level.",
        caption.above = TRUE)

```

Social Science Field: Social Sciences, Psychology & Cognitive Sciences, Economics & Business  

```{r}
#| label: tlb-model_social
#| tbl-cap: test
#| results: asis


social <- scholars %>% filter(field=="Social Sciences" | field=="Psychology & Cognitive Sciences" | field=="Economics & Business" )

model_base_soc <- lmer(h_index ~ 1+(1 | discipline), data=social)

# main effects are group mean centered

model_soc_disc <- lmer(h_index ~ 
                        I(prop_female-disc_prop_female)+I(100*disc_prop_female)+
                        I(100*(prop_sole-disc_prop_sole))+I(100*disc_prop_sole)+
                        (1|discipline), data=social)

# add a model with controls for age, specialization and the number of elite scholars per university (all grand mean centered)
model_soc_controls <- update(model_soc_disc, .~.+I(age-mean(age))+I(specialization-mean(specialization))+I(uni_pub_cnt-mean(uni_pub_cnt)))

knitreg(list(model_base_soc, model_soc_disc, model_soc_controls),
        custom.coef.names = c("Intercept",
                              "Probability [0-1] of female name (discipline mean centered)",
                              "Discipline mean percent female name",
                              "Percent sole authored publications (discipline mean centered)",
                              "Discipline mean percent sole authored publications",
                              "Age (grand mean centered)",
                              "Specialization (grand mean centered)",
                              "Highly productive university count (grand mean centered)"),
        custom.model.names = c("ANOVA Model","W-B Model","w/controls"),
        caption = "Multilevel models predicting H-index score for Social Science Field. Clustering at the disciplinary level.",
        caption.above = TRUE)

```

Non-Medical STEM Field: Physics & Astronomy, Engineering, Biology, Chemistry, Earth & Environmental Sciences, Mathematics & Statistics, Information & Communication Technologies  

```{r}
#| label: tlb-model_stem
#| tbl-cap: test
#| results: asis


stem <- scholars %>% filter(field=="Physics & Astronomy" | field=="Engineering" | field=="Biology" | field=="Chemistry" | field=="Earth & Environmental Sciences" | field == "Mathematics & Statistics" | field=="Information & Communication Technologies")

model_base_stem <- lmer(h_index ~ 1+(1 | discipline), data=stem)

# main effects are group mean centered

model_stem_disc <- lmer(h_index ~ 
                        I(prop_female-disc_prop_female)+I(100*disc_prop_female)+
                        I(100*(prop_sole-disc_prop_sole))+I(100*disc_prop_sole)+
                        (1|discipline), data=stem)

# add a model with controls for age, specialization and the number of elite scholars per university (all grand mean centered)
model_stem_controls <- update(model_stem_disc, .~.+I(age-mean(age))+I(specialization-mean(specialization))+I(uni_pub_cnt-mean(uni_pub_cnt)))

knitreg(list(model_base_stem, model_stem_disc, model_stem_controls),
        custom.coef.names = c("Intercept",
                              "Probability [0-1] of female name (discipline mean centered)",
                              "Discipline mean percent female name",
                              "Percent sole authored publications (discipline mean centered)",
                              "Discipline mean percent sole authored publications",
                              "Age (grand mean centered)",
                              "Specialization (grand mean centered)",
                              "Highly productive university count (grand mean centered)"),
        custom.model.names = c("ANOVA Model","W-B Model","w/controls"),
        caption = "Multilevel models predicting H-index score for STEM Field. Clustering at the disciplinary level.",
        caption.above = TRUE)

```

Humanaties Field: Communication & Textual Studies, Historical Studies, Philosophy & Theology, Visual & Performing Arts  


```{r}
#| label: tlb-model_human
#| tbl-cap: test
#| results: asis


human <- scholars %>% filter(field=="Communication & Textual Studies" | field=="Historical Studies" | field=="Philosophy & Theology" | field=="Visual & Performing Arts")

model_base_human <- lmer(h_index ~ 1+(1 | discipline), data=human)

# main effects are group mean centered

model_human_disc <- lmer(h_index ~ 
                        I(prop_female-disc_prop_female)+I(100*disc_prop_female)+
                        I(100*(prop_sole-disc_prop_sole))+I(100*disc_prop_sole)+
                        (1|discipline), data=human)

# add a model with controls for age, specialization and the number of elite scholars per university (all grand mean centered)
model_human_controls <- update(model_human_disc, .~.+I(age-mean(age))+I(specialization-mean(specialization))+I(uni_pub_cnt-mean(uni_pub_cnt)))

knitreg(list(model_base_human, model_human_disc, model_human_controls),
        custom.coef.names = c("Intercept",
                              "Probability [0-1] of female name (discipline mean centered)",
                              "Discipline mean percent female name",
                              "Percent sole authored publications (discipline mean centered)",
                              "Discipline mean percent sole authored publications",
                              "Age (grand mean centered)",
                              "Specialization (grand mean centered)",
                              "Highly productive university count (grand mean centered)"),
        custom.model.names = c("ANOVA Model","W-B Model","w/controls"),
        caption = "Multilevel models predicting H-index score for humanaties Field. Clustering at the disciplinary level.",
        caption.above = TRUE)

```